DeepPicar: A Low-cost Deep Neural Network-based Autonomous Car

### 2018-02-02 paper/background.tex
Minor changes to paper

To solve the problem of autonomous driving, a standard approach has
been decomposing the problem into multiple sub-problems,
such as lane marking detection, path planning, and low-level
control, which together form a processing pipeline~\cite{Bojarski2016}.
Recently, researchers are exploring another approach that dramatically
simplifies the standard control pipeline by applying deep neural
networks to directly produce control outputs from senor
inputs~\cite{Levine2016}. Figure~\ref{fig:end-to-end-control}
shows the differences between the two approaches.

@@

\subsection{Embedded Computing Platforms for Real-Time Inferencing}
%% \fixme{we can consider moving this section to a dedicated related work
%%   section at the end.}
Real-time embedded systems, such as an autonomous vehicle, present
unique challenges for deep learning, as the computing platforms of such
systems must satisfy two often conflicting goals~\cite{Otterness2017}:
%% Recent successes in AI, including NVIDIA's DAVE-2 showing, are due
%% in large part to the increased computing performance,
%% which afforded researchers to train and use ever deeper neural networks with
%% high accuracy.
%% For practical applications, the computer platform in an
%% autonomous vehicle must satisfy two often conflicting goals:
The platform must provide 
enough computing capacity for real-time processing of computationally
expensive AI workloads (deep neural networks);
The platform must also satisfy various
constraints such as cost, size, weight, and power consumption limits.

>>>>

To solve the problem of autonomous driving, a standard approach has
been decomposing the problem into multiple sub-problems,
such as lane marking detection, path planning, and low-level
control, which together form a processing pipeline~\cite{Bojarski2016}.
Recently, researchers are exploring another approach that dramatically
simplifies the standard control pipeline by applying deep neural
networks to directly produce control outputs from sensor
inputs~\cite{Levine2016}. Figure~\ref{fig:end-to-end-control}
shows the differences between the two approaches.

@@

\subsection{Embedded Computing Platforms for Real-Time Inferencing}
%% \fixme{we can consider moving this section to a dedicated related work
%%   section at the end.}
Real-time embedded systems, such as an autonomous vehicle, present
unique challenges for deep learning, as the computing platforms of such
systems must satisfy two often conflicting goals~\cite{Otterness2017}:
%% Recent successes in AI, including NVIDIA's DAVE-2 showing, are due
%% in large part to the increased computing performance,
%% which afforded researchers to train and use ever deeper neural networks with
%% high accuracy.
%% For practical applications, the computer platform in an
%% autonomous vehicle must satisfy two often conflicting goals:
(1) The platform must provide 
enough computing capacity for real-time processing of computationally
expensive AI workloads (deep neural networks); and
(2) The platform must also satisfy various
constraints such as cost, size, weight, and power consumption limits.

### 2018-02-02 paper/comparison.tex
Minor changes to paper

In this section, we compare three computing platforms---the Raspberry
Pi 3, the Intel UP~\footnote{http://www.up-board.org/up/} and NVIDIA
Jetson
TX2~\footnote{http://www.nvidia.com/object/embedded-systems-dev-kits-modules.html}---from
the point of view of supporting end-to-end deep learning
based autonomous vehicles. 
Table~\ref{tbl:platforms} shows architectural features of the three
platforms~\footnote{The GPU of Intel UP and the two Denver cores in the
  Tegra TX2 are not used in evaluation due to TensorFlow issues.}.

Our basic approach is to use the same DeepPicar software, and repeat
the experiments in Section~\ref{sec:evaluation} on each hardware
platform and compare the results. We do not, however, repeat the cache
partitioned experiments done on the Pi 3, as we did not believe our 
findings warranted further experimentation on additional platforms. 
For the Jetson TX2, we have two different system configurations,
which differ in whether TensorFlow is configured to use its GPU or
only the CPU cores. Thus, the total four system configurations are
compared.

@@

Finally, we compare the effect of co-scheduling memory bandwidth
intensive synthetic benchmarks on the DNN control loop timing. 
Figure~\ref{fig:sys_bench} shows the results.
As discussed in 
Section~\ref{sec:eval-memhog}, we observed dramatic execution time
increases, up to 11.6X, in Raspberry Pi 3 as we increased the number of
co-scheduled tasks. We also observe increased control loop execution
timing in the Intel Up and Jetson TX2, but the degree of the increase is 
not as dramatic as the Pi 3. Compared to their respective solo timings 
(i.e., the model runs on a single core in isolation), Intel UP suffers up to
2.3X execution time increase; TX2 (CPU) and TX2 (GPU) suffer up to
2.2X and 2.1X increases, respectively. This is somewhat suprising
because the Raspberry Pi 3's cores are in-order architecture based while
the cores in the Intel Up and NVIDIA TX2 are out-of-order architecture
based, and that the memory intensive tasks on out-of-order cores can
generate more memory traffic. We suspect that this is because the
memory subsystems in the Intel UP and TX2 platforms provide higher
bandwidth and fairness than the memory subsystem of the Pi 3. However, 
the L2 miss rates of the Pi 3 and TX2 platforms follow the same expected 
pattern, as can be seen in Figure~\ref{fig:sys_l2miss}. Namely, the 
percentage of L2 miss rates increases are more 
memory intensive co-runners are introduced. This was not the case for 
the UP board, where the model running on two cores generated the highest
percentage of L2 cache misses. This is most likely due to the CPU 
architecture of the UP's CPU which is Intel based, whereas both the Pi 3
and TX2 have ARM based CPUs.

>>>>

In this section, we compare three computing platforms---the Raspberry
Pi 3, the Intel UP~\footnote{http://www.up-board.org/up/} and the NVIDIA
Jetson
TX2~\footnote{http://www.nvidia.com/object/embedded-systems-dev-kits-modules.html}---from
the point of view of supporting end-to-end deep learning
based autonomous vehicles. 
Table~\ref{tbl:platforms} shows architectural features of the three
platforms~\footnote{The GPU of Intel UP and the two Denver cores in the
  Tegra TX2 are not used in evaluation due to TensorFlow issues.}.

Our basic approach is to use the same DeepPicar software, and repeat
the experiments in Section~\ref{sec:evaluation} on each hardware
platform and compare the results. 
%We do not, however, repeat the cache
%partitioned experiments done on the Pi 3, as we did not believe our 
%findings warranted further experimentation on additional platforms. 
For the Jetson TX2, we have two different system configurations,
which differ in whether TensorFlow is configured to use its GPU or
only the CPU cores. Thus, a total of four system configurations are
compared.

@@

Finally, we compare the effect of co-scheduling memory bandwidth
intensive synthetic benchmarks on the DNN control loop timing. 
Figure~\ref{fig:sys_bench} shows the results.
As discussed in 
Section~\ref{sec:eval-memhog}, we observed dramatic execution time
increases, up to 11.5X, in Raspberry Pi 3 as we increased the number of
co-scheduled tasks. We also observe increased control loop execution
timing in the Intel Up and Jetson TX2, but the degree of the increase is 
not as dramatic as the Pi 3. Compared to their respective solo timings 
(i.e., the model runs on a single core in isolation), Intel UP suffers up to
2.3X execution time increase; TX2 (CPU) and TX2 (GPU) suffer up to
2.2X and 2.1X increases, respectively. This is somewhat suprising
because the Raspberry Pi 3's cores are in-order architecture based while
the cores in the Intel Up and NVIDIA TX2 are out-of-order architecture
based, and that the memory intensive tasks on out-of-order cores can
generate more memory traffic. We suspect that this is because the
memory subsystems in the Intel UP and TX2 platforms provide higher
bandwidth and fairness than the memory subsystem of the Pi 3. However, 
the L2 miss rates of the Pi 3 and TX2 platforms follow the same expected 
pattern, as can be seen in Figure~\ref{fig:sys_l2miss}. Namely, the 
percentage of L2 miss rates increases are more 
memory intensive co-runners are introduced. This was not the case for 
the UP board, where the model running on two cores generated the highest
percentage of L2 cache misses. This is most likely due to the CPU 
architecture of the UP's CPU which is Intel based, whereas both the Pi 3
and TX2 have ARM based CPUs.

### 2018-02-02 paper/evaluation.tex
Minor changes to paper

Figure~\ref{fig:} shows the normalized execution time and L2 miss-rate
of the DNN model running on the Core 0 as a function of the number of
co-scheduled memory intensive synthetic benchmarks. First, as we
increase the number of co-runners, the DNN model's execution times are
increased---by up to 11.6X---even though the DNN model is running on a
dedicated core (Core 0). On the other hand, the DNN model's L2
cache-miss rates do not increase as much.
In other words, the DNN model's execution increase is not fully
explained by increased contention in the L2 cache space.
Instead, the high memory bandwidth pressure from the co-scheduled
memory intensive benchmarks is likely the primary cause.

@@

% \input{evaluation-cachepart}

@@

Fourth, we find that consolidating memory (DRAM) performance
intensive applications could jeopadize DNN performance, because DNN
performance appears to be very sensitive to memory performance; we 
observe up to 11.6X slowdown in DNN performance by co-scheduling 
synthetic memory bandwidth intensive applications on idle cores.

>>>>

Figure~\ref{fig:} shows the normalized execution time and L2 miss-rate
of the DNN model running on the Core 0 as a function of the number of
co-scheduled memory intensive synthetic benchmarks. First, as we
increase the number of co-runners, the DNN model's execution times are
increased---by up to 11.5X---even though the DNN model is running on a
dedicated core (Core 0). On the other hand, the DNN model's L2
cache-miss rates do not increase as much.
In other words, the DNN model's execution increase is not fully
explained by increased contention in the L2 cache space.
Instead, the high memory bandwidth pressure from the co-scheduled
memory intensive benchmarks is likely the primary cause.

@@

%% \input{evaluation-cachepart}

@@

Fourth, we find that consolidating memory (DRAM) performance
intensive applications could jeopadize DNN performance, because DNN
performance appears to be very sensitive to memory performance; we 
observe up to 11.5X slowdown in DNN performance by co-scheduling 
synthetic memory bandwidth intensive applications on idle cores.

### 2018-02-02 paper/intro.tex
Minor changes to paper

% how we trained (maybe moved to a later section)
We apply a standard imitation learning methodology to train the car to
follow tracks on the ground. We collect data for
training and validation by manually
controlling the RC car and recording the vision (from the webcam
mounted on the RC-car) and the human control inputs. We then train the
network offline using the collected data on a desktop computer, which
is equipped with a NVIDIA GTX 1060 GPU. Finally, the trained network is copied
back to the Raspberry Pi, which is then used to perform inference
operations---locally on the Pi---in the car's main control loop in
real-time. For real-time control, each inference operation must
be completed within the desired control period (e.g., 33.$\overline{\mbox{33}}$ ms
 period for 30 Hz control frequency).
% how we evaluated (in terms real-time performance)

@@

% performance on low-cost embedded multicore
Second, we find that real-time processing of DeepPicar's CNN-based
control loop is feasible on today's embedded computing platforms, even 
as inexpensive as the Raspberry Pi 3. We find that the control loop 
completes in under 33.$\overline{\mbox{33}}$ ms, or 30 Hz, using just two 
cores of the Raspberry Pi 3's quad-core CPU (w/o using its GPU), and can
do so 100\% of the time; or we can achieve 20 Hz control frequency using
just one core of the Raspberry Pi 3's quad-core CPU.
Other tested embedded platforms, Intel UP and NVIDIA TX2, offer even
better performance, and are capable of supporting deep-learning based 
real-time control over 100 Hz control frequency on the TX2 when its
embedded GPU is used.
We find that all the tested embedded computing platforms have 
enough computing capacity to consolidate multiple instances of
DeepPicar's DNN workload.

% all platforms have enough capacity to consolidate
Third, contention in shared hardware resources remains an important
issue when consolidating multiple workloads. 
In particular, we find that DNN inferencing is highly sensitive to
memory performance interference as we observe up to an 11.6X control
loop execution time increase, when memory performance intensive
applications are co-scheduled on idle CPU cores.
On the other hand, we find that DNN inferencing is not sensitive to
cache space.
%% ; we show that applying cache partitioning has negligible impact
%% in providing timing isolation on consolidated scenarios.
This finding
suggests that in order to guarantee real-time performance of DNN-based
control applications, solutions to guarantee their memory performance
are more important.

% contributions
The {\bf contributions} of this paper are as follows:
\begin{itemize}
  \item We present DeepPicar, a low-cost autonomous car testbed, which
    is based on a state-of-the-art end-to-end deep learning, as 
    open-source~\footnote{The source code, build 
instruction the DeepPicar
can be found at: \url{https://github.com/heechul/picar}}, with the
    goal of lowering the economic and safety-related barriers to study
    autonomous cars.

>>>>

% how we trained (maybe moved to a later section)
We apply a standard imitation learning methodology to train the car to
follow tracks on the ground. We collect data for
training and validation by manually
controlling the RC car and recording the vision (from the webcam
mounted on the RC-car) and the human control inputs. We then train the
network offline using the collected data on a desktop computer, which
is equipped with a NVIDIA GTX 1060 GPU. Finally, the trained network is copied
back to the Raspberry Pi 3, which is then used to perform inference
operations---locally on the Pi 3---in the car's main control loop in
real-time. For real-time control, each inference operation must
be completed within the desired control period (e.g., 33.$\overline{\mbox{33}}$ ms
 period for 30 Hz control frequency).
% how we evaluated (in terms real-time performance)

@@

% performance on low-cost embedded multicore
Second, we find that real-time processing of DeepPicar's CNN-based
control loop is feasible on today's embedded computing platforms, even 
as inexpensive as the Raspberry Pi 3. We find that the control loop 
completes in under 33.$\overline{\mbox{33}}$ ms, or 30 Hz, using just two 
cores of the Raspberry Pi 3's quad-core CPU (w/o using its GPU), and can
do so 100\% of the time; or we can achieve 20 Hz control frequency using
just one core of the Raspberry Pi 3's quad-core CPU.
Other tested embedded platforms, the Intel UP and NVIDIA TX2, offer even
better performance, and are capable of supporting deep-learning based 
real-time control over 100 Hz control frequency on the TX2 when its
embedded GPU is used.
We find that all the tested embedded computing platforms have 
enough computing capacity to consolidate multiple instances of
DeepPicar's DNN workload.

% all platforms have enough capacity to consolidate
Third, contention in shared hardware resources remains an important
issue when consolidating multiple workloads. 
In particular, we find that DNN inferencing is highly sensitive to
memory performance interference as we observe up to an 11.5X control
loop execution time increase, when memory performance intensive
applications are co-scheduled on idle CPU cores.
On the other hand, we find that DNN inferencing is not sensitive to
cache space.
%% ; we show that applying cache partitioning has negligible impact
%% in providing timing isolation on consolidated scenarios.
This finding
suggests that in order to guarantee real-time performance of DNN-based
control applications, solutions to guarantee their memory performance
are more important.

% contributions
The {\bf contributions} of this paper are as follows:
\begin{itemize}
  \item We present DeepPicar, a low-cost autonomous car testbed, which
    is based on a state-of-the-art end-to-end deep learning, as 
    open-source~\footnote{The source code, and build 
instructions for the DeepPicar
can be found at: \url{https://github.com/heechul/picar}}, with the
    goal of lowering the economic and safety-related barriers to study
    autonomous cars.

### 2018-02-02 paper/overview.tex
intro revision
Minor changes to paper

\section{DeepPicar Overview}

@@

In this section, we provide an overview of our DeepPicar platform.
In developing DeepPicar, one of our primary goals is to faithfully
replicate NVIDIA's DAVE-2 system on a smaller scale with using a low cost
multicore platform, the Raspberry Pi 3. Because the Raspberry Pi 3's
computing performance is much lower than that of the DRIVE PX platform
used in DAVE-2, we are interested in if, and how, we can process
computationally expensive neural network operations in
real-time. Specifically, inferencing (forward pass processing)
operations must be completed within each control period
duration---e.g., a WCET of 33.$\overline{\mbox{33}}$ ms for 30Hz control 
frequency---locally on the Pi 3 platform, although training of the 
network (back-propagation for weight updates) can be done offline and 
remotely using a desktop computer.
Figure~\ref{fig:overview} shows the DeepPicar, which is comprised of a
set of inexpensive components: a Raspberry Pi 3 Single Board Computer
(SBC), a Pololu DRV8835 motor driver, a Playstation Eye webcam, a
battery, and a 1:24 scale RC car. Table~\ref{tbl:carbom} shows the
estimated cost of the system.

@@

% data collection and training.
To collect the training data, a human pilot manually drives the RC car
on tracks we created (Figure~\ref{fig:track}) to record
timestamped videos and contol commands. The stored data is then copied 
to a desktop computer, which is equipped with a NVIDIA GTX 1050ti GPU, 
where we train the network to accelerate training speed. 
For comparison, training the network on the Raspbeery Pi 3 takes
approximately 4 hours, whereas it takes only about 4 minutes on the
desktop computer using the GTX 1050ti GPU.

@@

% inferencing on pi3
Once the network is trained on the desktop computer, the trained model
is copied back to the Raspberry Pi. The network is then used
by the car's main controller, which feeds a image frame from the web
camera as input to the network. In each control period, the produced 
steering angle output is then converted as the PWM values of the 
steering motor of the car. Figure~\ref{fig:controlloop} shows simplified 
pseudo code of the controller's main loop. Among the five steps, the 3rd step, 
network inferencing, is the most computationally intensive and dominates the
execution time.

@@

Other issues we observed in training/testing the network, which
can affect the prediction accuracy of the CNN, are camera and actuator
(motor) control latencies. In DeepPicar's
context, the camera latency is from the time the camera
sensor observes the scene to the time the computer actually reads the
digitized image data. This time can be noticable depending on the camera 
used and the performance of the Pi. Higher camera latency could
negatively affect control performance, because the DNN would analyze
stale scenes. Actuator (motor) control latency also can be a
factor, because moving the sterring motor to a desired position takes
time. 
In our platform, the combined latency is measured to be around
50 ms. In other words, in the recorded video and control data, a
control action (left, center, right) taken by the CNN appears to be
applied 50 ms later. Considering that camera's framerate is 30Hz
(33.3ms/frame), this is about two frames after the control action.
%% We experimentally measured the camera
%% latency and found it to be around 50-100 ms.
If this value is too high, control performance may suffer.
Our initial prototype suffered this problem as the observed camera and
control latency was as high as 300 ms, which negatively affected
control performance.
For reference, the latency of human perception is known to be as fast
as 13 ms~\cite{ThomasBurger2015}. 
% https://www.pubnub.com/blog/2015-02-09-how-fast-is-realtime-human-perception-and-technology/

Despite these issues, the trained models were still able to
achieve a reasonable degree of accuracy, successfully navigating
several different tracks we trained. In our testing, we found that the
DeepPicar could remain on a fairly challenging track for over 10
minutes at a moderate speed (50\% throttle), at which point we stop
the experiment, and more than one minute at a higher speed (75\%
throttle)~\footnote{A collection of self-driving videos can be found
  at: \url{https://photos.app.goo.gl/ce93sU7jPk4ywO8u2}}. Running at
higher speed is inherently more challenging because the CNN controller
has less time to recover from mistakes (bad predictions).  Also, we
find that the prediction accuracy is significantly affected by the
quality of training data as well as various environmental factors such
as the lighting condition. We plan to investigate more systematic ways
to improve the CNN's prediction accuracies.

>>>>

\section{DeepPicar Overview}\label{sec:overview}

@@

In this section, we provide an overview of our DeepPicar platform.
In developing DeepPicar, one of our primary goals is to faithfully
replicate NVIDIA's DAVE-2 system on a smaller scale using a low cost
multicore platform, the Raspberry Pi 3. Because the Raspberry Pi 3's
computing performance is much lower than that of the DRIVE PX platform
used in DAVE-2, we are interested in if, and how, we can process
computationally expensive neural network operations in
real-time. Specifically, inferencing (forward pass processing)
operations must be completed within each control period
duration---e.g., a WCET of 33.$\overline{\mbox{33}}$ ms for 30Hz control 
frequency---locally on the Pi 3 platform, although training of the 
network (back-propagation for weight updates) can be done offline and 
remotely using a desktop computer.
Figure~\ref{fig:overview} shows the DeepPicar, which is comprised of a
set of inexpensive components: a Raspberry Pi 3 Single Board Computer
(SBC), a Pololu DRV8835 motor driver, a Playstation Eye webcam, a
battery, and a 1:24 scale RC car. Table~\ref{tbl:carbom} shows the
estimated cost of the system.

@@

% data collection and training.
To collect the training data, a human pilot manually drives the RC car
on tracks we created (Figure~\ref{fig:tracks}) to record
timestamped videos and contol commands. The stored data is then copied 
to a desktop computer, which is equipped with a NVIDIA GTX 1050ti GPU, 
where we train the network to accelerate training speed. 
For comparison, training the network on the Raspbeery Pi 3 takes
approximately 4 hours, whereas it takes only about 4 minutes on the
desktop computer using the GTX 1050ti GPU.

@@

% inferencing on pi3
Once the network is trained on the desktop computer, the trained model
is copied back to the Raspberry Pi 3. The network is then used
by the car's main controller, which feeds a image frame from the web
camera as input to the network. In each control period, the produced 
steering angle output is then converted as the PWM values of the 
steering motor of the car. Figure~\ref{fig:controlloop} shows simplified 
pseudo code of the controller's main loop. Among the five steps, the 3rd step, 
network inferencing, is the most computationally intensive and dominates the
execution time.

@@

Other issues we observed in training/testing the network, which
can affect the prediction accuracy of the CNN, are camera and actuator
(motor) control latencies. In DeepPicar's
context, the camera latency is from the time the camera
sensor observes the scene to the time the computer actually reads the
digitized image data. This time can be noticable depending on the camera 
used and the performance of the Pi. Higher camera latency could
negatively affect control performance, because the DNN would analyze
stale scenes. Actuator (motor) control latency also can be a
factor, because moving the steering motor to a desired position takes
time. 
In our platform, the combined latency is measured to be around
50 ms. In other words, in the recorded video and control data, a
control action (left, center, right) taken by the CNN appears to be
applied 50 ms later. Considering that camera's framerate is 30Hz
(33.$\overline{\mbox{33}}$ ms/frame), this is about two frames after the 
control action.
%% We experimentally measured the camera
%% latency and found it to be around 50-100 ms.
If this value is too high, control performance may suffer.
Our initial prototype suffered this problem as the observed camera and
control latency was as high as 300 ms, which negatively affected
control performance.
For reference, the latency of human perception is known to be as fast
as 13 ms~\cite{ThomasBurger2015}. 
% https://www.pubnub.com/blog/2015-02-09-how-fast-is-realtime-human-perception-and-technology/

Despite these issues, the trained models were still able to
achieve a reasonable degree of accuracy, successfully navigating
several different tracks we used for training. In our testing, we found 
that the DeepPicar could remain on a fairly challenging track for over 
10 minutes at a moderate speed (50\% throttle), at which point we stopped
the experiment, and more than one minute at a higher speed (75\%
throttle)~\footnote{A collection of self-driving videos can be found
  at: \url{https://photos.app.goo.gl/ce93sU7jPk4ywO8u2}}. Running at
higher speed is inherently more challenging because the CNN controller
has less time to recover from mistakes (bad predictions).  Also, we
find that the prediction accuracy is significantly affected by the
quality of training data as well as various environmental factors such
as the lighting condition. We plan to investigate more systematic ways
to improve the CNN's prediction accuracies.
